Model Recognition as Planning

### 2018-11-02 doc/icaps-19/model-recognition.tex
all comments by Miquel processed

% Commented for blind submission
\author{Diego Aineto\and Sergio Jim\'enez\and Eva Onaindia \and Miquel Ram\'irez\\
{\small Departamento de Sistemas Inform\'aticos y Computaci\'on}\\
{\small Universitat Polit\`ecnica de Val\`encia.}\\
{\small Camino de Vera s/n. 46022 Valencia, Spain}\\
{\small \{dieaigar,serjice,onaindia\}@dsic.upv.es}}

\maketitle
\begin{abstract} 
Given a partially observed plan execution, and a set of possible planning models (models that share the same state variables but that update these variables according to different action models); {\em model recognition} is the task of identifying which model in the set produced/explains the given observation. The paper formalizes the {\em model recognition} task and introduces a novel method to estimate the probability of a \strips\ model to produce an observation of a plan execution. This method, that we called {\em model recognition as planning}, is built on top of off-the-shelf classical planning algorithms and is robust to missing intermediate states and actions in the observation of the plan execution. The effectiveness of {\em model recognition as planning} is shown in a set of \strips\ models encoding different kinds of {\em automata}. We show that {\em model recognition as planning} succeeds to identify the executed automata despite some state variables (e.g. the internal automata state) or the actual applied transitions are unobserved.
\end{abstract}

@@

\begin{figure}
  \begin{scriptsize}    
  \begin{center}
  \begin{tikzpicture}[scale=.4]
          \begin{scope}
            \draw (0, 0) grid (5, 5);
            \node[anchor=center] at (3.5, 1.5) {R};
            \node[anchor=center] at (0.5, 0.5) {};
            \draw[thick,style=dotted] (0.5,0.5) -- (4.5,0.5);
            \draw[thick,style=dotted] (4.5,0.5) -- (4.5,1.5);
            \draw[thick,style=dotted,->] (4.5,1.5) -- (3.7,1.5);                        
          \end{scope}
        \end{tikzpicture}
\hspace*{1cm}        
	\begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto,node distance=2cm,semithick]
	  \node[state] (A)              {$q_0$};
	  \node[state] (B) [right of=A] {$q_1$};          
	  \path
(A) edge [bend left, align=center] node {inc-y,\\dec-y} (B)
(B) edge [bend left, align=center] node {inc-y,\\dec-y} (A);
	\end{tikzpicture}
  \end{center}
  \end{scriptsize}  
 \caption{\small ({\em Left)} Robot navigating a $5\times 5$ grid. {\em (Right)} Automata to control that the robot only increments its x-coordinate when $q_0$ holds (actions {\tt inc-y} and {\tt dec-y} update the robot y-coordinate and switch the automata state).}
\label{fig:grid-example}
\end{figure}

To better illustrate {\em model recognition}, imagine a robot in a $n\times n$ grid whose navigation is determined by the \strips\ model of Figure~\ref{fig:model-example}. According to this model, the robot can increment its {\em x-coordinate} when {\tt\small $q0$} holds (i.e. at {\em even rows} if {\tt\small $q0$} holds initially) and decrement it when {\tt\small $q1$} holds (at {\em odd rows} if {\tt\small $q0$} holds initially). Apart from this particular navigation model, different action models could be defined within the same state variables (e.g. altering the way {\tt\small $q0$} and {\tt\small $q1$} are required and updated) and these models can determine different kinds of robot navigation. Given an observation of a plan execution (like the one illustrated at Figure~\ref{fig:grid-example}) {\em model recognition} aims to identify the action model that produced/explains that observation, despite relevant information is unobserved (e.g. the applied actions or the value of state variables {\small\tt $q0$} and {\small\tt $q1$}).

@@

The paper also introduces {\em model recognition as planning}; a novel method to estimate the probability of a given \strips\ model to produce an observed plan execution. Our method is built on top of off-the-shelf classical planning algorithms and is robust to missing intermediate states and actions in the observed plan execution. We evaluate the effectiveness of {\em model recognition as planning} with sets of \strips\ models that represent different {\em automata}. All the {\em automata} in a set are defined within the same state variables but different {\em transition functions}. We show that {\em model recognition as planning} succeeds to identify the executed {\em automata} despite some state variables (e.g. the internal automata state) or the actual applied transitions are unobserved.

@@

  (:action dec-y-odd
    :parameters (?v1 ?v2)
    :precondition (and (ycoord ?v1) (next ?v2 ?v1) (q1))
    :effect (and (not (ycoord ?v1)) (ycoord ?v2)
                 (not (q1)) (q0)))
  \end{verbatim}           
  \end{tiny}  
 \caption{\small Example of a \strips\ action model (given in PDDL) for robot navigation in a $n\times n$ grid.}
\label{fig:model-example}
\end{figure}

@@

A {\em classical planning frame} is a $\tup{F,A}$ pair, where $F$ is a set of {\em fluents} and $A$ is a set of {\em actions}. The semantics of actions $a\in A$ is specified with two functions: $\rho(s,a)$ that denotes whether the action $a$ is {\em applicable} in a state $s$ and $\theta(s,a)$ that denotes the {\em successor state} that results of applying the action $a$ in a state $s$. In this work we compactly represent the $\rho$ and $\theta$ functions following the \strips\ model. With this regard, an action $a\in A$ is defined by:
\begin{itemize}
\item $\pre(a)\in\mathcal{L}(F)$, the {\em preconditions} of $a$, is the set of literals that must hold for the action $a\in A$ to be applicable.
\item $\eff^+(a)\in\mathcal{L}(F)$, the {\em positive effects} of $a$, is the set of literals that are true after the application of action $a\in A$.
\item $\eff^-(a)\in\mathcal{L}(F)$, the {\em negative effects} of $a$, is the set of literals that are false after the application of the action.
\end{itemize}
We say that an action $a\in A$ is {\em applicable} in a state $s$ iff $\pre(a)\subseteq s$. The result of applying $a$ in $s$ is the {\em successor state} denoted by $\theta(s,a)=\{s\setminus\eff^-(a))\cup\eff^+(a)\}$.

A {\em classical planning problem} is a tuple $P=\tup{F,A,I,G}$, where $I$ is an initial state and $G\in\mathcal{L}(F)$ is a goal condition. A {\em plan} $\pi$ for $P$ is an action sequence $\pi=\tup{a_1, \ldots, a_n}$ that induces the {\em trajectory} $\tau(\pi,P)=\tup{a_1, s_1, \ldots, a_n, s_n}$ such that $s_0=I$ and, for each {\small $1\leq i\leq n$}, $a_i$ is applicable in $s_{i-1}$ and generates the successor state $s_i=\theta(s_{i-1},a_i)$. The {\em plan length} is denoted with $|\pi|=n$. A plan $\pi$ {\em solves} $P$ iff $G\subseteq s_n$, i.e.,~if the goal condition is satisfied at the last state reached after following the application of the plan $\pi$ in the initial state $I$. A solution plan for $P$ is {\em optimal} if it has minimum length.

@@

\subsection{The observation model}
Given a classical planning problem $P=\tup{F,A,I,G}$ and a trajectory $\tau(\pi,P)$ generated by the execution of a plan $\pi$ that solves $P$; the observation of the execution of $\pi$ on $P$ is $\mathcal{O}(\tau)=\tup{a_1^o,s_1^o \ldots , a_l^o, s_m^o}$, an interleaved combination of {\small $1\leq l\leq |\pi|$} observed actions and {\small $1\leq m\leq |\pi|$} observed states such that:
\begin{itemize}
\item Observed actions are {\em consistent} with $\pi$, which means that $\tup{a_1^o, \ldots, a_l^o}$ is a sub-sequence of $\pi$. 
\item Observed states are a sequence of {\em partial states} $\tup{s_1^o, \ldots, s_m^o}$ that is {\em consistent} with the sequence of states $\tup{s_0, s_1, \ldots, s_n}$ traversed by the trajectory $\tau(\pi,P)$. The initial state $I$ is fully observed while the observed states in $\mathcal{O}(\tau)$ may be partial states, i.e. the value of certain fluents in the intermediate states may not be observed (~$|s_i^o|\leq |F|$ for every $1\leq i\leq m$). Certain states may even be fully unobserved ($0\leq|s_i^o|$ for every $1\leq i\leq m$).  
\end{itemize}

With this regard, we are assuming that a bijective monotone mapping exists between plans and observations~\cite{ramirez2009plan}. We can then  conclude that transiting between two consecutive observed states in $\mathcal{O}(\tau)$ may require the execution of more than a single action ($\theta(s_i^o,\tup{a_1,\ldots,a_k})=s_{i+1}^o$, where ${\small k\geq 1}$ is unknown but finite). In other words, having $\mathcal{O}(\tau)$ does not imply knowing the actual length of $\pi$.

Figure~\ref{fig:grid-example} illustrates the {\em partial observation} of a six-state trajectory \{{\tt\scriptsize<(xcoord 2)(ycoord 1)>, <(xcoord 3)(ycoord 1)>, <(xcoord 4)(ycoord 1)>, <(xcoord 5)(ycoord 1)>, <(xcoord 5)(ycoord 2)>, <(xcoord 4)(ycoord 2)>}\}. In more detail, this observation only contains fluents of the kind {\tt\small (xcoord ?v)} and {\tt\small (ycoord ?v)} because the value of the remaining fluents, namely {\tt\small (next ?v1 ?v2)}, {\tt\small (q0)} and {\tt\small (q1)}, cannot be observed.

@@

\section{Model Recognition}
\label{sec:recognition}
The {\em model recognition} task is a tuple $\tup{P,M,\mathcal{O}}$ where:
\begin{itemize}
\item $P=\tup{F,A[\cdot],I,G}$ is a {\em classical planning problem} s.t. the semantics of actions $a\in A[\cdot]$ is unknown (i.e. for each $a\in A[\cdot]$ its corresponding $\rho$ and/or $\theta$ functions are undefined).
\item $M=\{\mathcal{M}_1,\ldots,\mathcal{M}_k\}$ is a non-empty {\em set of k models} for the actions in $A[\cdot]$ s.t., each model in $\mathcal{M}\in M$, defines different function pairs $\tup{\rho,\theta}$ over the state variables $F$.
\item $\mathcal{O}(\tau)$ is an {\em observation} of the execution of an unknown solution plan $\pi$ for the planning problem $P$.
\end{itemize}

{\em Model recognition} can be understood as a {\em classification task} where each class is represented with a different planning model $\mathcal{M}\in M$ and the observed plan execution $\mathcal{O}(\tau)$, is the single example to classify. The planning model associated to each class acts as the corresponding {\em class prototype} and it summarizes any observation of a plan execution that could be synthesized with that model (i.e. the set of all the examples that belong to that class).

We follow the {\em naive Bayes classifier} to assign a model $\mathcal{M}\in M$ to the given observation $\mathcal{O}(\tau)$. Therefore the {\em solution} to the {\em model recognition} task is the subset of models in $M$ that maximizes this expression.
\begin{align}
argmax_{\mathcal{M}\in M} P(\mathcal{O}|\mathcal{M}) P(\mathcal{M}).
\end{align}

\subsection{Formulating the $P(\mathcal{O}|\mathcal{M})$ {\em likelihood}}
The $P(\mathcal{M})$ probability expresses whether one model is known to be a priori more likely than the others. If this probability is not given as input it is reasonable to assume that, {\em a priori}, all models are equiprobable. The challenge in our formulation for {\em model recognition} is the definition of the $P(\mathcal{O}|\mathcal{M})$ {\em likelihood} that expresses the probability of observing $\mathcal{O}(\tau)$ when $\mathcal{M}$ is the planning model.

\begin{figure}
  \begin{scriptsize}
  \begin{center}
	\begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto,node distance=2cm,semithick]
	  \node[state] (A)              {$\mathcal{M}$};
	  \node[state] (B) [right of=A] {$\mathcal{M}'$};
	  \node[state] (C) [right of=B] {$\tau$};
	  \node[state] (D) [right of=C] {$\mathcal{O}$};          
	  \path
(A) edge  node {} (B)
(B) edge  node {} (C)
(C) edge  node {} (D)
;
	\end{tikzpicture}
  \end{center}
  \end{scriptsize}  
 \caption{\small {\em Bayesian network} abstracting that model $\mathcal{M}$ can be {\em transformed} into a model $\mathcal{M}'$ that is able to produce a trajectory $\tau(\pi,P)$ that reaches the goals in $P$ and is consistent with the observation $\mathcal{O}(\tau)$.}
\label{fig:net}
\end{figure}

Our approach to formulate the $P(\mathcal{O}|\mathcal{M})$ {\em likelihood} is to assess the cost of {\em transforming} $\mathcal{M}$ into a model $\mathcal{M'}$ such that: $\mathcal{M'}$ that can produce a trajectory $\tau(\pi,P)$ that reaches the goals in $P$ and is consistent with the observation $\mathcal{O}(\tau)$. Figure~\ref{fig:net} shows the four-variable {\em Bayesian network} abstracting this process. Regarding this network we have the following formulation:
\begin{align}
 P(\mathcal{O}|\mathcal{M})=\sum_{\mathcal{M}'}\sum_{\tau} P(\mathcal{M'}|\mathcal{M})P(\tau|\mathcal{M'})P(\mathcal{O}|\tau),
\end{align}
where $\tau$ ranges over all the trajectories that can be synthesized with a model $\mathcal{M}'$ and $\mathcal{M}'$ ranges over all the models that can be generated {\em transforming} $\mathcal{M}$.

Following this equation, the exact computation of $P(\mathcal{O}|\mathcal{M})$ is intractable. For most planning problems the set of trajectories consistent with an arbitrary observation can easily be huge (infinite in the case of planning problems without dead-ends)~\cite{lesh1995sound}. Even worse, the number of models $\mathcal{M}'$ that can be generated {\em transforming} a given classical planning model $\mathcal{M}$ explodes combinatorially. Instead, our approach is to formulate an effectively estimate of $P(\mathcal{O}|\mathcal{M})$ using the {\em edit distance} referred to \strips\ planning models. {\em Edit distances} are similarity metrics, traditionally computed over {\em strings} or {\em graphs}, and that has been proved successful for {\em pattern recognition}~\cite{masek1980faster,bunke1997relation}. In this work, we assess the cost of {\em transforming} a classical planning model $\mathcal{M}$ into a model $\mathcal{M'}$ formalizing and computing {\em edit distances} that are referred to \strips\ planning models.

\section{Recognition of \strips\ models}
\label{sec:asPlanning}
Here we analyze the particular instantiation of the {\em model recognition} task where the semantics of actions (i.e. $\rho$ and $\theta$ functions) are specified with \strips\ action schemata. We start formalizing the \strips\ schemata and define the full space of possible \strips\ schemata. Eventually, we introduce an {\em edit distance} for \strips\ schemata to estimate the $P(\mathcal{O}|\mathcal{M})$ likelihoods for classical planning models.

\subsection{Well-defined \strips\ action schemata}
\strips\ action schemata provide a compact representation for specifying classical planning models. Figure~\ref{fig:model-example} shows six \strips\ action schemata, codded in PDDL, that determine a particular kind of robot navigation in $n\times n$ grids.

{\em A \strips\ action schemata} $\xi$ is defined by a list of {\em parameters} $pars(\xi)$, and three list of predicates (namely $pre(\xi)$, $del(\xi)$ and $add(\xi)$) that shape the kind of fluents that can appear in the {\em preconditions}, {\em negative effects} and {\em positive effects} of the actions induced from that schemata.

We say that two \strips\ schemata $\xi$ and $\xi'$ are {\em comparable} iff $pars(\xi)=pars(\xi')$, both share the same list of parameters. For instance, we claim that the six action schemata of Figure~\ref{fig:model-example} are {\em comparable} while, for example, the {\small\tt stack(?v1,?v2)} and {\small\tt pickup(?v1)} schemata from the four operator {\em blocksworld}~\cite{slaney2001blocks} are not. Last but not least, we say that two \strips\ models $\mathcal{M}$ and $\mathcal{M}'$ are {\em comparable} iff there exists a bijective function $\mathcal{M} \mapsto \mathcal{M}^*$ that maps every action schemata $\xi\in\mathcal{M}$ to a comparable schemata $\xi'\in\mathcal{M'}$ and vice versa.

Given a \strips\ action schemata $\xi$ and a set of {\em predicates} $\Psi$ that shape the propositional state variables. The set of FOL interpretations of $\Psi$ over the corresponding parameters of the schemata $\xi$, confines the elements that can appear in the $pre(\xi)$, $del(\xi)$ and $add(\xi)$ lists. We denote this set of FOL interpretations as ${\mathcal I}_{\Psi,\xi}$. For any of the six action schemata defined in Figure~\ref{fig:model-example} the ${\mathcal I}_{\Psi,\xi}$ set contains the same ten elements, ${\mathcal I}_{\Psi,\xi}=${\small\tt\{xcoord($v_1$), xcoord($v_2$), ycoord($v_1$), ycoord($v_2$), q0(), q1(), next($v_1$,$v_1$), next($v_1$,$v_2$), next($v_2$,$v_1$), next($v_2$,$v_2$))\}}.

>>>>

% Commented for blind submission
\author{Diego Aineto \and Sergio Jim\'enez\and Eva Onaindia\\
{\scriptsize Departamento de Sistemas Inform\'aticos y Computaci\'on}\\
{\scriptsize Universitat Polit\`ecnica de Val\`encia.}\\
{\scriptsize Camino de Vera s/n. 46022 Valencia, Spain}\\
{\scriptsize \{dieaigar,serjice,onaindia\}@dsic.upv.es}
\And Miquel Ram\'irez\\
{\scriptsize School of Computing and Information Systems}\\
{\scriptsize The University of Melbourne}\\
{\scriptsize Melbourne, Victoria. Australia}\\
{\scriptsize miquel.ramirez@nimelb.edu.au}}

\maketitle
\begin{abstract} 
Given a partially observed plan execution, and a set of possible planning models (models that share the same state variables but that update these variables according to different action models); {\em model recognition} is the task of identifying which model in the set produced/explains the given observation. The paper formalizes the {\em model recognition} task and introduces a novel method to estimate the probability of a \strips\ model to produce an observation of a plan execution. This method, that we called {\em model recognition as planning}, is built on top of off-the-shelf classical planning algorithms and is robust to missing intermediate states and actions in the observation of the plan execution. The effectiveness of {\em model recognition as planning} is shown in a set of \strips\ models encoding different kinds of {\em automata}. We show that {\em model recognition as planning} succeeds to identify the executed automata despite the actual applied transitions, plus some state variables e.g. the internal automata state, are unobserved.
\end{abstract}

@@

\begin{figure}
  \begin{scriptsize}    
  \begin{center}
  \begin{tikzpicture}[scale=.4]
          \begin{scope}
            \draw (0, 0) grid (5, 5);
            \node[anchor=center] at (2.5, 1.5) {R};
            \node[anchor=center] at (0.5, 0.5) {};
            \draw[thick,style=dotted] (0.5,0.5) -- (3.5,0.5);
            \draw[thick,style=dotted] (3.5,0.5) -- (3.5,1.5);
            \draw[thick,style=dotted,->] (3.5,1.5) -- (2.7,1.5);                        
          \end{scope}
        \end{tikzpicture}
\hspace*{1cm}        
	\begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto,node distance=2cm,semithick]
	  \node[state] (A)              {$q_0$};
	  \node[state] (B) [right of=A] {$q_1$};          
	  \path
(A) edge [bend left, align=center] node {inc-y,\\dec-y} (B)
(B) edge [bend left, align=center] node {inc-y,\\dec-y} (A);
	\end{tikzpicture}
  \end{center}
  \end{scriptsize}  
 \caption{\small ({\em Left)} Robot navigating a $5\times 5$ grid. {\em (Right)} Automata to control that the robot only increments its x-coordinate when $q_0$ holds (actions {\tt inc-y} and {\tt dec-y} update the robot y-coordinate and switch the automata state).}
\label{fig:grid-example}
\end{figure}

To better illustrate {\em model recognition}, imagine a robot in a $N\times N$ grid whose navigation is determined by the \strips\ model of Figure~\ref{fig:model-example}. According to this model, the robot can increment its {\em x-coordinate} when {\tt\small $q0$} holds (i.e. at {\em even rows} if {\tt\small $q0$} holds initially) and decrement it when {\tt\small $q1$} holds (at {\em odd rows} if {\tt\small $q0$} holds initially). Apart from the model of Figure~\ref{fig:model-example}, different models can be defined within the same state variables (e.g. altering the way {\tt\small $q0$} and {\tt\small $q1$} are required and updated) and these models can shape different kinds of robot navigation. Given a partially observed plan execution (like the one illustrated at Figure~\ref{fig:grid-example} where state variables {\small\tt $q0$} and {\small\tt $q1$} are unobserved) {\em model recognition} aims to identify the action model that produced/explains that observation.

@@

The paper also introduces {\em model recognition as planning}; a novel method to estimate the probability of a given \strips\ model to produce an observed plan execution. Our method is built on top of off-the-shelf classical planning algorithms and is robust to missing intermediate states and actions in the observed plan execution. We evaluate the effectiveness of {\em model recognition as planning} with sets of \strips\ models that represent different {\em automata}. All the {\em automata} in a set are defined within the same state variables but with different {\em transition functions}. We show that {\em model recognition as planning} identifies the executed {\em automata} despite the actual applied transitions, plus some state variables (e.g. the internal automata state), are unobserved.

@@

  (:action dec-y-odd
    :parameters (?v1 ?v2)
    :precondition (and (ycoord ?v1) (next ?v2 ?v1) (q1))
    :effect (and (not (ycoord ?v1)) (ycoord ?v2)
                 (not (q1)) (q0)))
  \end{verbatim}           
  \end{tiny}  
 \caption{\small Example of a \strips\ action model (given in PDDL) for robot navigation in a $N\times N$ grid.}
\label{fig:model-example}
\end{figure}

@@

A {\em classical planning frame} is a $\tup{F,A}$ pair, where $F$ is a set of {\em fluents} and $A$ is a set of {\em actions}. The semantics of actions $a\in A$ is specified with two functions: $\rho(s,a)$ that denotes whether action $a$ is {\em applicable} in a state $s$ and $\theta(s,a)$ that denotes the {\em successor state} that results of applying action $a$ in a state $s$. In this work we compactly represent the $\rho$ and $\theta$ functions of an action $a\in A$ with the \strips\ model:
\begin{itemize}
\item $\pre(a)\in\mathcal{L}(F)$, the {\em preconditions} of $a$, is the set of literals that must hold for the action $a\in A$ to be applicable.
\item $\eff^+(a)\in\mathcal{L}(F)$, the {\em positive effects} of $a$, is the set of literals that are true after the application of action $a\in A$.
\item $\eff^-(a)\in\mathcal{L}(F)$, the {\em negative effects} of $a$, is the set of literals that are false after the application of the action.
\end{itemize}
We say that an action $a\in A$ is {\em applicable} in a state $s$ iff $\pre(a)\subseteq s$. The result of applying $a$ in $s$ is the {\em successor state} denoted by $\theta(s,a)=\{s\setminus\eff^-(a))\cup\eff^+(a)\}$.

A {\em classical planning problem} is a tuple $P=\tup{F,A,I,G}$, where $I$ is an initial state and $G\in\mathcal{L}(F)$ is a goal condition. A {\em plan} $\pi$ is an action sequence $\pi=\tup{a_1, \ldots, a_n}$ and $|\pi|=n$ denotes its {\em plan length}. The execution of a plan $\pi$ on a classical planning problem $P$ induces a {\em trajectory} $\tau(\pi,P)=\tup{a_1, s_1, \ldots, a_n, s_n}$ such that $s_0=I$ and, for each {\small $1\leq i\leq n$}, $a_i$ is applicable in $s_{i-1}$ and generates the successor state $s_i=\theta(s_{i-1},a_i)$. A plan $\pi$ {\em solves} $P$ iff $G\subseteq s_n$, i.e.,~if the goal condition is satisfied at the last state reached after following the execution of plan $\pi$ on problem $P$. A solution plan for $P$ is {\em optimal} iff it has minimum length.

@@

\subsection{The observation model}
Given a classical planning problem $P=\tup{F,A,I,G}$, a solution plan $\pi$ for $P$, and a trajectory $\tau(\pi,P)$ generated by the execution of $\pi$ on $P$; the observation of the execution of $\pi$ on $P$ is $\mathcal{O}(\tau)=\tup{a_1^o,s_1^o \ldots , a_l^o, s_m^o}$, an interleaved combination of {\small $1\leq l\leq |\pi|$} observed actions and {\small $1\leq m\leq |\pi|$} observed states:
\begin{itemize}
\item {\bf Observed actions} are {\em consistent} with $\pi$, which means that $\tup{a_1^o, \ldots, a_l^o}$ is a sub-sequence of $\pi$. 
\item {\bf Observed states} are a sequence of {\em partial states} $\tup{s_1^o, \ldots, s_m^o}$ that is {\em consistent} with the sequence of states $\tup{s_0, s_1, \ldots, s_n}$ traversed by the trajectory $\tau(\pi,P)$. The initial state $I$ is fully observed while $\tup{s_1^o, \ldots, s_m^o}$ may be partial states, i.e. the value of certain fluents in the intermediate states may not be observed (~$|s_i^o|\leq |F|$ for $1\leq i\leq m$). Even more, certain intermediate states may be fully unobserved ($0\leq|s_i^o|$ for $1\leq i\textless m$).  
\end{itemize}

We are assuming that a bijective monotone mapping exists between plans and observations~\cite{ramirez2009plan}. Therefore transiting between two consecutive observed states in $\mathcal{O}(\tau)$ may require the execution of more than a single action ($\theta(s_i^o,\tup{a_1,\ldots,a_k})=s_{i+1}^o$, where ${\small k\geq 1}$ is unknown but finite). In other words, having $\mathcal{O}(\tau)$ does not imply knowing the actual length of $\pi$.

Figure~\ref{fig:grid-example} illustrates the {\em partial observation} of a five-state trajectory \{{\tt\scriptsize<(xcoord 2)(ycoord 1)>, <(xcoord 3)(ycoord 1)>, <(xcoord 4)(ycoord 1)>, <(xcoord 4)(ycoord 2)>, <(xcoord 3)(ycoord 2)>}\}. In more detail, this observation only contains fluents of the kind {\tt\small (xcoord ?v)} and {\tt\small (ycoord ?v)}. The value of the remaining fluents, namely {\tt\small (next ?v1 ?v2)}, {\tt\small (q0)} and {\tt\small (q1)}, cannot be observed.

@@

\section{Model Recognition}
\label{sec:recognition}
The {\em model recognition} task is a tuple $\tup{P,M,\mathcal{O}}$ where:
\begin{itemize}
\item $P=\tup{F,A[\cdot],I,G}$ is a {\em classical planning problem} where $A[\cdot]$ is a set of actions s.t. the semantics of actions $a\in A[\cdot]$ is unknown (i.e. for each $a\in A[\cdot]$ its corresponding $\rho$ and/or $\theta$ functions are undefined).
\item $M=\{\mathcal{M}_1,\ldots,\mathcal{M}_k\}$ is a {\em set of k different models} for the actions in $A[\cdot]$ s.t., each model in $\mathcal{M}\in M$, defines different semantics for the actions in $A[\cdot]$ (i.e. different function pairs $\tup{\rho,\theta}$) but using the same set of state variables $F$.
\item $\mathcal{O}(\tau)$ is an {\em observation} of the execution of an unknown solution plan $\pi$ for the planning problem $P$.
\end{itemize}

{\em Model recognition} can be understood as a {\em classification task} where each class is represented with a different planning model $\mathcal{M}\in M$ and the observed plan execution $\mathcal{O}(\tau)$, is the single example to classify. The planning model associated to each class acts as the corresponding {\em class prototype} and it summarizes any observation of a plan execution that could be synthesized with that model (i.e. the set of all the examples that belong to that class). To assign a model $\mathcal{M}\in M$ to a given observation $\mathcal{O}(\tau)$, we follow the {\em naive Bayes classifier}. The {\em solution} to the {\em model recognition} task is then the subset of models in $M$ that maximizes this expression.
\begin{align}
argmax_{\mathcal{M}\in M} P(\mathcal{O}|\mathcal{M}) P(\mathcal{M}).
\end{align}

\subsection{Formulating the $P(\mathcal{O}|\mathcal{M})$ likelihood}
The $P(\mathcal{M})$ probability expresses whether one model is known to be a priori more likely than the others. If this probability is not given as input it is reasonable to assume that, {\em a priori}, all models are equiprobable. The challenge in our formulation for {\em model recognition} is the definition of the $P(\mathcal{O}|\mathcal{M})$ {\em likelihood} that expresses the probability of observing $\mathcal{O}(\tau)$ when $\mathcal{M}$ is the planning model.

\begin{figure}
  \begin{scriptsize}
  \begin{center}
	\begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto,node distance=2cm,semithick]
	  \node[state] (A)              {$\mathcal{M}$};
	  \node[state] (B) [right of=A] {$\mathcal{M}'$};
	  \node[state] (C) [right of=B] {$\tau$};
	  \node[state] (D) [right of=C] {$\mathcal{O}$};          
	  \path
(A) edge  node {} (B)
(B) edge  node {} (C)
(C) edge  node {} (D)
;
	\end{tikzpicture}
  \end{center}
  \end{scriptsize}  
 \caption{\small {\em Bayesian network} abstracting that model $\mathcal{M}$ can be {\em transformed} into a model $\mathcal{M}'$ that produces a trajectory $\tau(\pi,P)$ that (1), reaches the goals in $P$ and (2), is consistent with $\mathcal{O}(\tau)$.}
\label{fig:net}
\end{figure}

Our approach to formulate the $P(\mathcal{O}|\mathcal{M})$ {\em likelihood} is to assess the cost of {\em transforming} $\mathcal{M}$ into a model $\mathcal{M'}$ that produces a trajectory $\tau(\pi,P)$ such that: (1), $\tau(\pi,P)$ reaches the goals in $P$ and (2), $\tau(\pi,P)$ is consistent with observation $\mathcal{O}(\tau)$. Figure~\ref{fig:net} shows the {\em Bayesian network} that abstracts this procedure. Regarding this network we have the following formulation of the $P(\mathcal{O}|\mathcal{M})$ {\em likelihood}:
\begin{align}
 P(\mathcal{O}|\mathcal{M})=\sum_{\mathcal{M}'}\sum_{\tau} P(\mathcal{M'}|\mathcal{M})P(\tau|\mathcal{M'})P(\mathcal{O}|\tau),
\end{align}
where $\tau$ ranges over all the trajectories consistent with $\mathcal{O}(\tau)$ that can be synthesized with a model $\mathcal{M}'$ and $\mathcal{M}'$ ranges over all the models that can be generated {\em transforming} $\mathcal{M}$.

The exact computation of $P(\mathcal{O}|\mathcal{M})$ with equation (2) is intractable. For most planning problems the set of trajectories consistent with an arbitrary observation can easily be huge, infinite in the case of planning problems without dead-ends~\cite{lesh1995sound}. Even worse, the number of models $\mathcal{M}'$ that can be generated {\em transforming} a given classical planning model $\mathcal{M}$ explodes combinatorially with the number of state variables. Instead, our approach is to formulate an effectively estimate of $P(\mathcal{O}|\mathcal{M})$ using an {\em edit distance} referred to \strips\ planning models. {\em Edit distances} are similarity metrics, traditionally computed over {\em strings} or {\em graphs}, and that has been proved successful for {\em pattern recognition}~\cite{masek1980faster,bunke1997relation}. In this work, we assess the cost of {\em transforming} a classical planning model $\mathcal{M}$ into a model $\mathcal{M'}$ computing {\em edit distances} between \strips\ planning models.

\section{Recognition of \strips\ models}
\label{sec:asPlanning}
Here we study the particular instantiation of {\em model recognition} where the semantics of models $\mathcal{M}\in M$ (i.e. the corresponding $\rho$ and $\theta$ functions) is specified with \strips\ action schemata. We start formalizing \strips\ schemata and define the space of possible \strips\ schemata for a given set of state variables. Eventually, we introduce an {\em edit distance} for \strips\ schemata to estimate the $P(\mathcal{O}|\mathcal{M})$ likelihoods for classical planning models.

\subsection{Well-defined \strips\ action schemata}
\strips\ action schemata provide a compact representation for specifying classical planning models. Figure~\ref{fig:model-example} shows six \strips\ action schemata that shape a particular kind of robot navigation in $N\times N$ grids (no matter the grid size).

{\em A \strips\ action schemata} $\xi$ is defined by four lists: A list of {\em parameters} $pars(\xi)$, and three list of predicates (namely $pre(\xi)$, $del(\xi)$ and $add(\xi)$) that shape the kind of fluents that can appear in the {\em preconditions}, {\em negative effects} and {\em positive effects} of the actions induced from that schemata.

\begin{definition}[Comparable \strips\ action schemata]
Two \strips\ schemata $\xi$ and $\xi'$ are {\bf comparable} iff $pars(\xi)=pars(\xi')$, both share the same list of parameters.
\end{definition}
For instance, we claim that the six action schemata of Figure~\ref{fig:model-example} are {\em comparable} while, for example, the {\small\tt stack(?v1,?v2)} and {\small\tt pickup(?v1)} schemata from a four operator {\em blocksworld}~\cite{slaney2001blocks} are not. Last but not least, we say that two \strips\ models $\mathcal{M}$ and $\mathcal{M}'$ are {\em comparable} iff there exists a bijective function $\mathcal{M} \mapsto \mathcal{M}^*$ that maps every action schemata $\xi\in\mathcal{M}$ to a comparable schemata $\xi'\in\mathcal{M'}$ and vice versa.

Given the set of {\em predicates} $\Psi$, that shape the propositional state variables, and a list of {\em parameters} $pars(\xi)$. The set of elements that can appear in the $pre(\xi)$, $del(\xi)$ and $add(\xi)$ of the \strips\ action schemata $\xi$ is given by FOL interpretations of $\Psi$ over the parameters $pars(\xi)$. We denote this set of FOL interpretations as ${\mathcal I}_{\Psi,\xi}$. For any of the six action schemata defined in Figure~\ref{fig:model-example} the ${\mathcal I}_{\Psi,\xi}$ set contains the same ten elements, ${\mathcal I}_{\Psi,\xi}=${\small\tt\{xcoord($v_1$), xcoord($v_2$), ycoord($v_1$), ycoord($v_2$), q0(), q1(), next($v_1$,$v_1$), next($v_1$,$v_2$), next($v_2$,$v_1$), next($v_2$,$v_2$))\}}.

Despite any element from ${\mathcal I}_{\Psi,\xi}$ can {\em a priori} appear in the $pre(\xi)$, $del(\xi)$ and $add(\xi)$ of schemata $\xi$, the space of possible \strips\ schemata is constrained by a set ${\mathcal C}$ that includes: 
\begin{itemize}
\item {\em Syntactic constraints}. \strips\ constraints require $del(\xi)\subseteq pre(\xi)$, $del(\xi)\cap add(\xi)=\emptyset$ and $pre(\xi)\cap add(\xi)=\emptyset$. Considering exclusively these syntactic constraints, the size of the space of possible \strips\ schemata is given by $2^{2\times|{\mathcal I}_{\Psi,\xi}|}$. For every action schemata in the navigation model of Figure~\ref{fig:model-example} then $2^{2\times 10}=1,048,576$.
\item {\em Domain-specific constraints}. One can introduce domain-specific knowledge to constrain further the space of possible schemata. For instance in a {\em robot navigation} model like the one in Figure~\ref{fig:model-example}, {\small\tt q0()} and {\small\tt q1()} are exclusive so they cannot hold at the same time in a $pre(\xi)$/$del(\xi)$/$add(\xi)$ list. Further, {\small\tt next($v_1$,$v_1$)} and {\small\tt next($v_2$,$v_2$)} will not appear at any of these lists because the {\tt\small next} predicate codes the {\em successor} function for {\em natural numbers}. These domain-specific constraints reduce further the size of the space of possible action schemata to $2^{2\times 7}=16,384$ (for every schemata in the navigation model of Figure~\ref{fig:model-example}).
\end{itemize}
